---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```



# Survalis: Unified Survival Machine Learning and Interpretability in R

`survalis` provides a unified framework for survival machine learning survival analysis in R. It supports a wide range of learners, evaluation metrics, cross-validation and interpretability methods. 



## Installation

```{r, eval = FALSE}
# Install from GitHub
remotes::install_github("ielbadisy/survalis")

# Or from source
install.packages("survalis_0.6.0.tar.gz", repos = NULL, type = "source")
```



## Core philosophy

* Consistent function patterns: `fit_*()`, `predict_*()`, `tune_*()`
* Learners return standardized `mlsurv_model` objects
* Predictions return `data.frame` of survival probabilities: `t=100`, `t=200`, ...
* Evaluation is fully modular: plug any `fit_*`/`predict_*` with `cv_survlearner()` or `score_survmodel()`
* Designed for interpretability post prediction. 



## Exploring the package

### List all available survival learners

```{r}
library(survalis)
# See all available learners
list_survlearners()

# See only tunable learners (those with a tune_* function)
list_survlearners(has_tune = TRUE)

# Shortcut for tunable learners
list_tunable_survlearners()
```

### List interpretability tools

```{r}
# List available interpretability methods
list_interpretability_methods()

# Show which compute_* methods have a plot_* counterpart
subset(list_interpretability_methods(), !is.na(plot))
```

### List evaluation metrics

```{r}
# List available metrics used in cross-validation and scoring
list_metrics()
```



## Basic Workflow

**1. Fit a model**
```{r}
mod_cox <- fit_coxph(Surv(time, status) ~ age + karno + celltype, data = veteran)
summary(mod_cox)
```


**2. Predict survival probabilities**

```{r}
pred <- predict_coxph(mod_cox, newdata = veteran[1:5, ], times = c(100, 200))
pred
```


**3. Evaluate model performance**

Direct evalution (single split): 
```{r}
score <- score_survmodel(mod_cox, times = c(100, 200), metrics = c("cindex", "ibs"))
score
```


```{r}
cv_res <- cv_survlearner(
  Surv(time, status) ~ age + karno + celltype,
  veteran,
  fit_coxph,
  predict_coxph,
  times  = 80,
  metrics = c("cindex", "ibs"),
  folds = 5,
  seed = 123,
  verbose = FALSE
  )

cv_res
```

```{r}
cv_summary(cv_res)
```



**4. Visualize interpretation**
```{r}
shap_meanabs <- compute_shap(
  model         = mod_cox,
  newdata       = veteran[100,],
  baseline_data = veteran,
  times         = 80,
  sample.size   = 50,
  aggregate     = TRUE,
  method        = "meanabs"
  )

shap_meanabs
```

```{r}
plot_shap(shap_meanabs)
```

### More interpretability methods

`survalis` also provides PDP, ALE, surrogate explanations, tree surrogates,
permutation importance, interaction analysis, and counterfactuals.

**Partial dependence and ICE**

```{r}
pdp_age <- compute_pdp(
  model = mod_cox,
  data = veteran,
  feature = "age",
  times = c(100, 200, 300),
  method = "pdp+ice"
  )

plot_pdp(pdp_age, feature = "age", which = "per_time")
plot_pdp(pdp_age, feature = "age", which = "integrated", smooth = TRUE)
```

**Accumulated local effects**

```{r}
ale_karno <- compute_ale(
  model = mod_cox,
  newdata = veteran,
  feature = "karno",
  times = c(100, 200, 300)
  )

plot_ale(ale_karno, feature = "karno", which = "per_time")
plot_ale(ale_karno, feature = "karno", which = "integrated", smooth = TRUE)
```

**Local surrogate explanation**

```{r}
local_surrogate <- compute_surrogate(
  model = mod_cox,
  newdata = veteran[1, , drop = FALSE],
  baseline_data = veteran,
  times = c(100, 200, 300),
  target_time = 200,
  k = 5
  )

local_surrogate
plot_surrogate(local_surrogate, top_n = 10)
```

**Tree surrogate**

```{r}
tree_surrogate <- compute_tree_surrogate(
  model = mod_cox,
  data = veteran,
  times = c(100, 200, 300)
  )

plot_tree_surrogate(tree_surrogate, type = "importance", top_n = 5)
# plot_tree_surrogate(tree_surrogate, type = "tree")
```

**Permutation variable importance**

```{r}
varimp_res <- compute_varimp(
  model = mod_cox,
  times = c(100, 200, 300),
  metric = "ibs",
  n_repetitions = 5,
  seed = 123
  )

varimp_res
plot_varimp(varimp_res)
```

**Feature interactions**

```{r}
interaction_1way <- compute_interactions(
  model = mod_cox,
  data = veteran,
  times = c(100, 200, 300),
  target_time = 200,
  type = "1way"
  )

interaction_heatmap <- compute_interactions(
  model = mod_cox,
  data = veteran,
  times = c(100, 200, 300),
  target_time = 200,
  type = "heatmap"
  )

interaction_time <- compute_interactions(
  model = mod_cox,
  data = veteran,
  times = c(100, 200, 300),
  type = "time"
  )

plot_interactions(interaction_1way, type = "1way")
plot_interactions(interaction_heatmap, type = "heatmap")
plot_interactions(interaction_time, type = "time")
```

**Counterfactual explanations**

```{r}
counterfactuals <- compute_counterfactual(
  model = mod_cox,
  newdata = veteran[1, , drop = FALSE],
  times = c(100, 200, 300),
  target_time = 200,
  features_to_change = c("age", "karno", "diagtime"),
  cost_penalty = 0.01
  )

counterfactuals
```

**5. Calibration**

```{r}
compute_calibration(
   model = mod_cox, data = veteran,
   time = "time", status = "status",
   eval_time = 80, n_bins = 10, n_boot = 30
   ) |> plot_calibration()
```



## Citing

```r
citation("survalis")
```
