---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```



# Survalis: Unified Survival Modeling and Interpretability in R

`survalis` provides a standardized and modular framework for machine learning survival analysis in R. It supports a wide range of learners, evaluation metrics, cross-validation and interpretability methods. 



## Installation

```{r, eval = FALSE}
# Install from GitHub
remotes::install_github("ielbadisy/survalis")

# Or from source
install.packages("survalis_0.6.0.tar.gz", repos = NULL, type = "source")
```



## Core philosophy

* Consistent function patterns: `fit_*()`, `predict_*()`, `tune_*()`
* Learners return standardized `mlsurv_model` objects
* Predictions return `data.frame` of survival probabilities: `t=100`, `t=200`, ...
* Evaluation is fully modular: plug any `fit_*`/`predict_*` with `cv_survlearner()` or `score_survmodel()`
* Designed for interpretability post prediction. 



## Exploring the package

### List all available survival learners

```{r}
library(survalis)
# See all available learners
list_survlearners()

# See only tunable learners (those with a tune_* function)
list_survlearners(has_tune = TRUE)

# Shortcut for tunable learners
list_tunable_survlearners()
```

### List interpretability tools

```{r}
# List available interpretability methods
list_interpretability_methods()

# Show which compute_* methods have a plot_* counterpart
subset(list_interpretability_methods(), !is.na(plot))
```

### List evaluation metrics

```{r}
# List available metrics used in cross-validation and scoring
list_metrics()
```



## Basic Workflow

**1. Fit a model**
```{r}
mod_cox <- fit_coxph(Surv(time, status) ~ age + karno + celltype, data = veteran)
summary(mod_cox)
```


**2. Predict survival probabilities**

```{r}
pred <- predict_coxph(mod_cox, newdata = veteran[1:5, ], times = c(100, 200))
pred
```


**3. Evaluate model performance**

Direct evalution (single split): 
```{r}
score <- score_survmodel(mod_cox, times = c(100, 200), metrics = c("cindex", "ibs"))
score
```


```{r}
cv_res <- cv_survlearner(
  Surv(time, status) ~ age + karno + celltype,
  veteran,
  fit_coxph,
  predict_coxph,
  times  = 80,
  metrics = c("cindex", "ibs"),
  folds = 5,
  seed = 123,
  verbose = FALSE
  )

cv_res
```

```{r}
cv_summary(cv_res)
```



**4. Visualize interpretation**
```{r}
shap_meanabs <- compute_shap(
  model         = mod_cox,
  newdata       = veteran[100,],
  baseline_data = veteran,
  times         = 80,
  sample.size   = 50,
  aggregate     = TRUE,
  method        = "meanabs"
  )

shap_meanabs
```

```{r}
plot_shap(shap_meanabs)
```


**5. Calibration**

```{r}
compute_calibration(
   model = mod_cox, data = veteran,
   time = "time", status = "status",
   eval_time = 80, n_bins = 10, n_boot = 30
   ) |> plot_calibration()
```



## Citing

```r
citation("survalis")
```


