---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```



# Survalis: Unified Survival Machine Learning and Interpretability in R

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/survalis)](https://CRAN.R-project.org/package=survalis)
[![R-CMD-check](https://github.com/ielbadisy/survalis/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ielbadisy/survalis/actions/workflows/R-CMD-check.yaml)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/grand-total/survalis)](https://CRAN.R-project.org/package=survalis)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

`survalis` provides a unified framework for survival machine learning in R. It
supports 22 learners, evaluation metrics, cross-validation, benchmarking, and
model-agnostic interpretability methods through a common survival-probability
matrix (`survmat`) interface.

`survalis` is available on CRAN: <https://CRAN.R-project.org/package=survalis>.

## New in version 1.3.0: the seven-verb interface

```{r, eval = FALSE}
library(survalis)
f <- Surv(time, status) ~ age + karno + celltype

m   <- fit(f, veteran, model = "ranger", spec = list(num.trees = 500))
predict(m, veteran[1:5, ], times = c(90, 180, 365))          # survmat
evaluate(m, times = c(90, 180, 365), resampling = cv(5))     # resampled metrics

cmp <- compare(f, veteran, models = c("coxph", "ranger", "survdnn"),
               times = c(90, 180, 365))
plot(cmp)

plot(interpret(m, "shap", newdata = veteran[1, ], times = 180))
estimate(f, veteran, model = "ranger", treatment = "trt", tau = 365)
```

`fit()` / `predict()` / `evaluate()` / `tune()` / `compare()` / `interpret()` /
`estimate()` cover the whole workflow. The granular `fit_*()` / `predict_*()` /
`compute_*()` / `benchmark*()` functions remain fully available; see
`vignette("from-1.0", package = "survalis")` and `?`survalis-deprecated``.



## Installation

```{r, eval = FALSE}
# Install the released version from CRAN
install.packages("survalis")

# Or install the development version from GitHub
remotes::install_github("ielbadisy/survalis")
```



## New in version 1.1.0

Version 1.1.0 adds a set of downstream tools that work with the same fitted
models and `survmat` predictions used throughout the package:

* `screen_fdr()` performs univariate Cox screening with Benjamini-Hochberg FDR
  control and an optional minimum-screen guarantee.
* `estimate_rmst()` and `plot_rmst()` estimate standardized marginal RMST or a
  two-group RMST contrast with percentile-bootstrap intervals.
* `compute_shap_matrix()` and `plot_shap_beeswarm()` summarize time-integrated
  SHAP contributions across observations.
* `roc_survmat()` and `dca_survmat()` provide IPCW time-dependent ROC and
  decision-curve analyses, with `plot_roc()` and `plot_dca()` helpers.

```{r, eval = FALSE}
# FDR screening
selected <- screen_fdr(
  Surv(time, status) ~ age + karno + diagtime + prior,
  data = veteran,
  alpha = 0.1,
  minscreen = 2
)

# Standardized RMST contrast between treatment groups
rmst <- estimate_rmst(
  mod_cox,
  data = veteran,
  tau = 200,
  times = seq(10, 300, by = 10),
  trt_col = "trt",
  R = 200,
  seed = 1
)

# Time-dependent ROC and decision curves from survmat predictions
pred_100 <- predict_coxph(mod_cox, veteran, times = 100)
roc <- roc_survmat(Surv(veteran$time, veteran$status), pred_100, t_star = 100)
dca <- dca_survmat(Surv(veteran$time, veteran$status), pred_100, t_star = 100)
plot_roc(roc)
plot_dca(dca)
```



## Core philosophy

* Seven verbs cover the whole workflow: `fit()`, `predict()`, `evaluate()`, `tune()`, `compare()`, `interpret()`, `estimate()`
* Every learner returns a standardized `survalis_fit` (a `mlsurv_model` subclass)
* Matrix predictions return a `survmat`: columns `t=100`, `t=200`, ..., with the grid recorded in `attr(x, "times")`
* Evaluation is fully modular: any learner id works with `evaluate()`/`compare()` under any `cv()`/`holdout()`/`group_cv()`/`bootstrap()` resampling
* Designed for interpretability post prediction, through a single `interpret()` entry point



## Exploring the package

### List all available survival learners

```{r}
library(survalis)
# See all available learners
list_survlearners()

# See only tunable learners (those with a tune_* function)
list_survlearners(has_tune = TRUE)

# Shortcut for tunable learners
list_tunable_survlearners()
```

### List interpretability tools

```{r}
# List available interpretability methods
list_interpretability_methods()

# Show which compute_* methods have a plot_* counterpart
subset(list_interpretability_methods(), !is.na(plot))
```

### List evaluation metrics

```{r}
# List available metrics used in cross-validation and scoring
list_metrics()
```



## Basic Workflow

**1. Fit a model**
```{r}
mod_cox <- fit(Surv(time, status) ~ age + karno + celltype, veteran, model = "coxph")
mod_cox
```


**2. Predict survival probabilities**

```{r}
pred <- predict(mod_cox, newdata = veteran[1:5, ], times = c(100, 200))
pred
```

`predict()` also returns other quantities via `type`: `"risk"` (1 - S(t)),
`"chf"`, `"hazard"`, `"rmst"`, `"quantile"`, `"median"`.


**3. Evaluate model performance**

```{r}
ev <- evaluate(mod_cox, times = c(80, 160), resampling = cv(5, seed = 123))
ev
```

```{r}
summary(ev)
```


**4. Compare multiple learners**

`compare()` evaluates every learner on the *same* resampling folds, so results
are paired; `tune = TRUE` tunes each tunable learner first.

```{r}
cmp <- compare(
  Surv(time, status) ~ age + karno + celltype,
  data = veteran,
  models = c("coxph", "rpart", "ranger"),
  times = c(80, 160),
  resampling = cv(3, seed = 1)
  )

cmp
plot(cmp)
```



**5. Kaplan-Meier curves**

`plot_survcurve()` produces a styled Kaplan-Meier curve with a confidence
ribbon, log-rank p-value, and an aligned number-at-risk table, in the spirit
of `survminer::ggsurvplot()` but implemented natively (no dependency on
survminer).

```{r}
plot_survcurve(Surv(time, status) ~ trt, data = veteran)
```



**6. Visualize interpretation**

`interpret()` is the single entry point to every explanation method; its
`plot()` dispatches to the matching `plot_*()` helper.

```{r}
shap_meanabs <- interpret(
  mod_cox, "shap",
  newdata     = veteran[100, ],
  times       = 80,
  sample.size = 50,
  aggregate   = TRUE,
  shap_method = "meanabs"
  )

as.data.frame(shap_meanabs)
```

```{r}
plot(shap_meanabs)
```

### More interpretability methods

`survalis` also provides PDP, ALE, surrogate explanations, tree surrogates,
permutation importance, interaction analysis, and counterfactuals -- all
through `interpret(fit, method = ..., ...)`.

**Partial dependence and ICE**

```{r}
pdp_age <- interpret(mod_cox, "pdp", feature = "age", times = c(100, 200, 300))

plot(pdp_age, which = "per_time")
plot(pdp_age, which = "integrated", smooth = TRUE)
```

**Accumulated local effects**

```{r}
ale_karno <- interpret(mod_cox, "ale", feature = "karno", times = c(100, 200, 300))

plot(ale_karno, which = "per_time")
plot(ale_karno, which = "integrated", smooth = TRUE)
```

**Local surrogate explanation**

```{r}
local_surrogate <- interpret(
  mod_cox, "surrogate",
  newdata     = veteran[1, , drop = FALSE],
  times       = c(100, 200, 300),
  target_time = 200,
  k           = 5
  )

as.data.frame(local_surrogate)
plot(local_surrogate, top_n = 10)
```

**Tree surrogate**

```{r}
tree_surrogate <- interpret(mod_cox, "tree_surrogate", times = c(100, 200, 300))

plot(tree_surrogate, type = "importance", top_n = 5)
# plot(tree_surrogate, type = "tree")
```

**Permutation variable importance**

```{r}
varimp_res <- interpret(
  mod_cox, "varimp",
  times          = c(100, 200, 300),
  metric         = "ibs",
  n_repetitions  = 5,
  seed           = 123
  )

as.data.frame(varimp_res)
plot(varimp_res)
```

**Feature interactions**

```{r}
interaction_1way <- interpret(mod_cox, "interaction", times = c(100, 200, 300),
                              target_time = 200, type = "1way")
interaction_heatmap <- interpret(mod_cox, "interaction", times = c(100, 200, 300),
                                 target_time = 200, type = "heatmap")
interaction_time <- interpret(mod_cox, "interaction", times = c(100, 200, 300),
                              type = "time")

plot(interaction_1way, type = "1way")
plot(interaction_heatmap, type = "heatmap")
plot(interaction_time, type = "time")
```

**Counterfactual explanations**

```{r}
counterfactuals <- interpret(
  mod_cox, "counterfactual",
  newdata            = veteran[1, , drop = FALSE],
  times              = c(100, 200, 300),
  target_time        = 200,
  features_to_change = c("age", "karno", "diagtime"),
  cost_penalty       = 0.01
  )

as.data.frame(counterfactuals)
```

**7. Calibration**

```{r}
interpret(mod_cox, "calibration", eval_time = 80, n_bins = 10, n_boot = 30) |> plot()
```



## Citing

```r
citation("survalis")
```
